% LaTeX source for textbook ``How to think like a computer scientist''
% Copyright (c)  2001  Allen B. Downey, Jeffrey Elkner, and John Dewey.

% Permission is granted to copy, distribute and/or modify this
% document under the terms of the GNU Free Documentation License,
% Version 1.1  or any later version published by the Free Software
% Foundation; with the Invariant Sections being "Contributor List",
% with no Front-Cover Texts, and with no Back-Cover Texts. A copy of
% the license is included in the section entitled "GNU Free
% Documentation License".

% This distribution includes a file named fdl.tex that contains the text
% of the GNU Free Documentation License.  If it is missing, you can obtain
% it from www.gnu.org or by writing to the Free Software Foundation,
% Inc., 59 Temple Place - Suite 330, Boston, MA 02111-1307, USA.

\chapter{Contributor List}

To paraphrase the philosophy of the Free Software Foundation, this
book is free like free speech, but not necessarily free like free
pizza.  It came about because of a collaboration that would not have
been possible without the GNU Free Documentation License.  So we
thank the Free Software Foundation for developing this license
and, of course, making it available to us.

We also thank the more than 100 sharp-eyed and
thoughtful readers who have sent us suggestions and corrections over
the past few years.  In the spirit of free software, we decided to
express our gratitude in the form of a contributor list.  Unfortunately,
this list is not complete, but we are doing our best to keep it
up to date.

If you have a chance to look through the list, you should
realize that each person here has spared you and all subsequent
readers from the confusion of a technical error or a
less-than-transparent explanation, just by sending us a note.

Impossible as it may seem after so many corrections, there may still
be errors in this book.  If you should stumble across one, please
check the online version of the book at {\tt http://thinkpython.com},
which is the most up-to-date version.  If the error has not been
corrected, please take a minute to send us email at {\tt
feedback@thinkpython.com}.  If we make a change due to your
suggestion, you will appear in the next version of the contributor
list (unless you ask to be omitted).  Thank you!


\begin{itemize}

\item Lloyd Hugh Allen sent in a correction to Section 8.4.
%He can be reached at: {\tt lha2@columbia.edu}

\item Yvon Boulianne sent in a correction of a semantic error in
Chapter 5.
%She can be reached at: {\tt mystic@monuniverse.net}

\item Fred Bremmer submitted a correction in Section 2.1.
%He can be reached at:  {\tt Fred.Bremmer@ubc.cu}

\item Jonah Cohen wrote the Perl scripts to convert the
LaTeX source for this book into beautiful HTML.

%His Web page is {\tt jonah.ticalc.org}
%and his email is {\tt JonahCohen@aol.com}

\item Michael Conlon sent in a grammar correction in Chapter 2
and an improvement in style in Chapter 1, and he initiated discussion
on the technical aspects of interpreters.

%Michael can be reached at: {\tt michael.conlon@sru.edu}

\item Benoit Girard sent in a
correction to a humorous mistake in Section 5.6.

%Benoit can be reached at:
%{\tt benoit.girard@gouv.qc.ca}

\item Courtney Gleason and Katherine Smith wrote {\tt horsebet.py},
which was used as a case study in an earlier version of the book.  Their
program can now be found on the website.

%Courtney can be reached at: {\tt
%orion1558@aol.com}

\item Lee Harr submitted more corrections than we have room to list
here, and indeed he should be listed as one of the principal editors
of the text.

%He can be reached at: {\tt missive@linuxfreemail.com}

\item James Kaylin is a student using the text. He has submitted
numerous corrections.

%James can be reached by email at: {\tt Jamarf@aol.com}

\item David Kershaw fixed the broken {\tt catTwice} function in Section
3.10.

%He can be reached at: {\tt david\_kershaw@merck.com}

\item Eddie Lam has sent in numerous corrections to Chapters 
1, 2, and 3.
He also fixed the Makefile so that it creates an index the first time it is
run and helped us set up a versioning scheme.  

%Eddie can be reached at:
%{\tt nautilus@yoyo.cc.monash.edu.au}

\item Man-Yong Lee sent in a correction to the example code in
Section 2.4.  

%He can be reaced at: {\tt yong@linuxkorea.co.kr}

\item David Mayo pointed out that the word ``unconsciously"
in Chapter 1 needed
to be changed to ``subconsciously".

%David can be reached at:{\tt bdbear44@netscape.net}

\item Chris McAloon sent in several corrections to Sections 3.9 and
3.10.

%He can be reached at: {\tt cmcaloon@ou.edu}

\item Matthew J. Moelter has been a long-time contributor who sent
in numerous corrections and suggestions to the book.  

%He can be reached at:
%{\tt mmoelter@calpoly.edu}

\item Simon Dicon Montford reported a missing function definition and
several typos in Chapter 3.  He also found errors in the {\tt increment}
function in Chapter 13.

%He can be reached at: {\tt dicon@bigfoot.com}

\item John Ouzts corrected the definition of ``return value"
in Chapter 3.

%He can be reached at: {\tt jouzts@bigfoot.com}

\item Kevin Parks sent in valuable comments and suggestions as to how
to improve the distribution of the book.

%He can be reached at: {\tt cpsoct@lycos.com}

\item David Pool sent in a typo in the glossary of Chapter 1, as well
as kind words of encouragement.

%He can be reached at: {\tt pooldavid@hotmail.com}

\item Michael Schmitt sent in a correction to the chapter on files
and exceptions.

%He can be reached at: {\tt ipv6\_128@yahoo.com}

\item Robin Shaw pointed out an error in Section 13.1, where the
printTime function was used in an example without being defined.

%Robin can be reached at: {\tt randj@iowatelecom.net}

\item Paul Sleigh found an error in Chapter 7 and a bug in Jonah Cohen's
Perl script that generates HTML from LaTeX.

%He can be reached at: {\tt bat@atdot.dotat.org}

%\item Christopher Smith is a computer science teacher at the Blake
%School in Minnesota who teaches Python to his beginning students.

%He can be reached at: {\tt csmith@blakeschool.org or smiles@saysomething.com}

\item Craig T. Snydal is testing the text in a course at Drew
University.  He has contributed several valuable suggestions and corrections.

%and can be reached at: {\tt csnydal@drew.edu}

\item Ian Thomas and his students are using the text in a programming
course.  They are the first ones to test the chapters in the latter half
of the book, and they have made numerous corrections and suggestions.

%Ian can be reached at: {\tt ithomas@sd70.bc.ca}

\item Keith Verheyden sent in a correction in Chapter 3.

%He can be reached at: {\tt kverheyd@glam.ac.uk}

\item Peter Winstanley let us know about a longstanding error in
our Latin in Chapter 3.

%He can be reached at:{\tt Peter.Winstanley@scotland.gsi.gov.uk} 

\item Chris Wrobel made corrections to the code in the chapter on
file I/O and exceptions. 

%He can be reached at: {\tt ferz980@yahoo.com}

\item Moshe Zadka has made invaluable contributions to this project.
In addition to writing the first draft of the chapter on Dictionaries, he
provided continual guidance in the early stages of the book.

%He can be reached at: {\tt moshez@math.huji.ac.il}

\item Christoph Zwerschke sent several corrections and
pedagogic suggestions, and explained the difference between {\em gleich}
and {\em selbe}.

\item James Mayer sent us a whole slew of spelling and
typographical errors, including two in the contributor list.

% james.mayer@acm.org

\item Hayden McAfee caught a potentially confusing inconsistency
between two examples.
%hayden.mcafee@mindspring.com

\item Angel Arnal is part of an international team of translators
working on the Spanish version of the text.  He has also found several
errors in the English version.

\item Tauhidul Hoque and Lex Berezhny created the illustrations
in Chapter 1 and improved many of the other illustrations.

\item Dr. Michele Alzetta caught an error in Chapter 8 and sent
some interesting pedagogic comments and suggestions about Fibonacci
and Old Maid.
%mikalzet@libero.it

\item Andy Mitchell caught a typo in Chapter 1 and a broken example
in Chapter 2.
%phantom917@hotmail.com

\item Kalin Harvey suggested a clarification in Chapter 7 and
caught some typos.
%kalin@metamuscle.net

\item Christopher P. Smith caught several typos and is helping us
prepare to update the book for Python 2.2.
%csmith@blakeschool.org

\item David Hutchins caught a typo in the Foreword.
%jsdah2@uas.alaska.edu

\item Gregor Lingl is teaching Python at a high school in Vienna,
Austria.  He is working on a German translation of the book,
and he caught a couple of bad errors in Chapter 5.
%glingl@aon.at

%Sean McShane sent us a very nice note
%sean.mcshane@sheridanc.on.ca

\item Julie Peters caught a typo in the Preface.
%jkpeters@dmacc.cc.ia.us

\item Florin Oprina sent in an improvement in {\tt makeTime},
a correction in {\tt printTime}, and a nice typo.
%oprina@student.uit.no 

\item D.~J.~Webre suggested a clarification in Chapter 3.
%d_webre@yahoo.com

% \item 
% jkane@broadlink.com

\item Ken found a fistful of errors in Chapters 8, 9 and 11.
%ken@codeweavers.com

\item Ivo Wever caught a typo in Chapter 5 and suggested a clarification
in Chapter 3.
% I.J.W.Wever@student.tnw.tudelft.nl

% rbeumer@knijnenberg.nl

\item Curtis Yanko suggested a clarification in Chapter 2.
% YankoC@gspinc.com

\item Ben Logan sent in a number of typos and problems with translating
the book into HTML.
%ben@wblogan.net

%\item XXX suggested a clarification in Chapter 7, but prefers not
% to be included here.
%ejykfy@comcast.net

%\item Florian Thiel caught an inconsistency in Chapter 2.
%noroute@web.de

\item Jason Armstrong saw the missing word in Chapter 2.
%jarmstrong@technicacorp.com

\item Louis Cordier noticed a spot in Chapter 16 where the code
didn't match the text.
% lcordier@dsp.sun.ac.za

\item Brian Cain suggested several clarifications in Chapters 2 and 3.
% Brian.Cain@motorola.com

\item Rob Black sent in a passel of corrections, including some
changes for Python 2.2.
% Rob.Black@static2358.com

\item Jean-Philippe Rey at Ecole Centrale
Paris sent a number of patches, including some updates for Python 2.2
and other thoughtful improvements.
%<jean-philippe.rey@ecp.fr>

\item Jason Mader at George Washington University made a number
of useful suggestions and corrections.
%Jason Mader <jason@ncac.gwu.edu>

\item Jan Gundtofte-Bruun reminded us that ``a error'' is an error.
% Jan Gundtofte-Bruun <jan@g-b.dk>

\item Abel David and Alexis Dinno reminded us that the plural of
``matrix'' is ``matrices'', not ``matrixes''.  This error was in the
book for years, but two readers with the same initials reported it on
the same day.  Weird.
% Abel David <abel.david@gmail.com>, lexy-lou@doyenne.com

\item Charles Thayer encouraged us to get rid of the semi-colons
we had put at the ends of some statements and to clean up our
use of ``argument'' and ``parameter''.
% Charles Thayer <catintp@yahoo.com>

\item Roger Sperberg pointed out a twisted piece of logic in Chapter 3.
%<rsperberg@gmail.com>

\item Sam Bull pointed out a confusing paragraph in Chapter 2.
%Sam Bull <dreamsorcerer@gmail.com>

\item Andrew Cheung pointed out two instances of ``use before def.''
%cheunga@u.washington.edu

\item Hans Batra found an error in Chapter 16.

\item Chris Seberino suggested some improvements in the Preface.

\item Yuri Takhteyev pointed out a problem with single and double quotes.

\end{itemize}


% correspondents

% Python version

% Sam
% "sgasster@muon.com" <sgasster@muon.com>
% suggestions about syntax glossary

% James Pomeroy
% moodykre8r@earthlink.net
% suggestion about PEMDAS

% Joel Jensen  << joel@ens.net >>
% translated the book into MS html


% Gregg Boggs <boggs+@pitt.edu>
% correction to Pythagorean theorem

% John P. Withers
% jp_withers@yahoo.com
% question about latex

% Michael Brownfield
% mmc81@airmail.net
% general kind words

% wolfgang teschner
% wtr@hannover.sgh-net.de
% general kind words

% Jeff
% jca@po.cwru.edu
% problem with pdf, and nice comments

% d_webre@yahoo.com
% confusing comment about calculation of pi

% Michael Wheatfill
% mwheatfill@tse-us.com
% general good comments and question about histograms

% Java version

% Vladimir
% "pisemsky@pisem.net"
% general praise

% Yong
% ybakos69@yahoo.com" <ybakos69@yahoo.com>
% clarification of interface and abstract class





